sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-03-11 doc/paper-jss/inla-comparison.Rmd
Make sections work with TeX #13

# Comparison between \pkg{INLA}, \pkg{inlabru}, and \pkg{sdmTMB} {#app:inla}

@@

## \pkg{INLA} version

@@

## \pkg{inlabru} version

@@

## \pkg{sdmTMB} version

@@

```{r maps, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \pkg{sdmTMB} model.", fig.align='center'}
inverse_link <- m$family$linkinv
ggplot(pred) +
  geom_raster(aes(X, Y, fill = inverse_link(est))) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

@@

```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \pkg{inlabru} model.", fig.align='center'}
nd_bru <- nd
nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
sp::coordinates(nd_bru) <- c("X", "Y")
nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))

>>>>

\section[Comparison between INLA, inlabru, and sdmTMB]{Comparison between \pkg{INLA}, \pkg{inlabru}, and \pkg{sdmTMB}}\label{app:inla}

@@

\subsection[INLA version]{\pkg{INLA} version}

@@

\subsection[inlabru version]{\pkg{inlabru} version}

@@

\subsection[sdmTMB version]{\pkg{sdmTMB} version}

@@

```{r maps, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \\pkg{sdmTMB} model.", fig.align='center'}
inverse_link <- m$family$linkinv
ggplot(pred) +
  geom_raster(aes(X, Y, fill = inverse_link(est))) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

@@

```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \\pkg{inlabru} model.", fig.align='center'}
nd_bru <- nd
nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
sp::coordinates(nd_bru) <- c("X", "Y")
nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))

### 2023-03-11 doc/paper-jss/paper-jss.Rmd
Make sections work with TeX #13

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2022].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [e.g., @heaton2019].
The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that relies on \proglang{R}-\pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].

@@

```{r child="model-description.Rmd", eval=FALSE}
```

>>>>

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2022].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [e.g., @heaton2019].
The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that relies on \proglang{R}-\pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].

@@

### 2023-03-11 doc/paper-jss/timing.Rmd
Make sections work with TeX #13

## Illustration of generating the INLA mesh

@@

## Example sdmTMB model fit

@@

## Example spaMM model fit

@@

## Example inlabru model fit

@@

## Example mgcv model fit

>>>>

## Illustration of generating the \textbf{INLA} mesh

@@

\subsection[Example sdmTMB model fit]{Example \pkg{sdmTMB} model fit}

@@

\subsection[Example spaMM model fit]{Example \pkg{spaMM} model fit}

@@

\subsection[Example inlabru model fit]{Example \pkg{inlabru} model fit}

@@

\subsection[Example mgcv model fit]{Example \pkg{mgcv} model fit}
